HW 8 

###Exercise 1

1) 

```{r}
#| warning: false
#| message: false

library(vegawidget)

supply_demand <- read.csv("swd-lets-practice-ex-2-3.csv")

supply_demand$capacity <- as.numeric(gsub(",", "", supply_demand$capacity))
supply_demand$demand <- as.numeric(gsub(",", "", supply_demand$demand))

long_data <- data.frame(
  date = rep(supply_demand$date, 2),
  type = c(rep("CAPACITY", nrow(supply_demand)), rep("DEMAND", nrow(supply_demand))),
  value = c(supply_demand$capacity, supply_demand$demand)
)

spec_1a <- list(
  `$schema` = vega_schema(),
  data = list(values = long_data),
  mark = "bar",
  width = 500,
  height = 300,
  encoding = list(
    x = list(
      field = "date",
      type = "nominal",
      axis = list(title = NULL, labelAngle = -45)
    ),
    y = list(
      field = "value",
      type = "quantitative",
      axis = list(title = NULL)
    ),
    xOffset = list(
      field = "type",
      type = "nominal"
    ),
    color = list(
      field = "type",
      type = "nominal",
      scale = list(
        domain = list("DEMAND", "CAPACITY"),
        range = list("#6baed6", "#bdd7e7")
      ),
      legend = list(title = NULL, orient = "top-right")
    )
  )
)

vegawidget(as_vegaspec(spec_1a))
```

2)
```{r}
#| warning: false
#| message: false

spec_1b <- list(
  `$schema` = vega_schema(),
  data = list(values = long_data),
  mark = list(type = "line"),
  width = 500,
  height = 300,
  encoding = list(
    x = list(
      field = "date",
      type = "nominal",
      axis = list(title = NULL, labelAngle = -45)
    ),
    y = list(
      field = "value",
      type = "quantitative",
      axis = list(title = NULL)
    ),
    color = list(
      field = "type",
      type = "nominal",
      scale = list(
        domain = list("DEMAND", "CAPACITY"),
        range = list("#6baed6", "#bdd7e7")
      ),
      legend = list(title = NULL, orient = "top-right")
    )
  )
)

vegawidget(as_vegaspec(spec_1b))
```

3)
```{r}
#| warning: false
#| message: false

spec_1c <- list(
  `$schema` = vega_schema(),
  data = list(values = long_data),
  mark = list(type = "bar", opacity = 0.7),
  width = 500,
  height = 300,
  encoding = list(
    x = list(
      field = "date",
      type = "nominal",
      axis = list(title = NULL, labelAngle = -45)
    ),
    y = list(
      field = "value",
      type = "quantitative",
      axis = list(title = NULL)
    ),
    color = list(
      field = "type",
      type = "nominal",
      scale = list(
        domain = list("DEMAND", "CAPACITY"),
        range = list("#6baed6", "#bdd7e7")
      ),
      legend = list(title = NULL, orient = "top-right")
    )
  )
)

vegawidget(as_vegaspec(spec_1c))
```

4)
```{r}
#| warning: false
#| message: false

supply_demand$unmet <- supply_demand$demand - supply_demand$capacity

unmet_data <- data.frame(
  date = rep(supply_demand$date, 2),
  type = c(rep("CAPACITY", nrow(supply_demand)), rep("UNMET DEMAND", nrow(supply_demand))),
  value = c(supply_demand$capacity, supply_demand$unmet)
)

spec_1d <- list(
  `$schema` = vega_schema(),
  data = list(values = unmet_data),
  mark = list(type = "bar"),
  width = 500,
  height = 300,
  encoding = list(
    x = list(
      field = "date",
      type = "nominal",
      axis = list(title = NULL, labelAngle = -45)
    ),
    y = list(
      field = "value",
      type = "quantitative",
      stack = TRUE,
      axis = list(title = NULL)
    ),
    color = list(
      field = "type",
      type = "nominal",
      scale = list(
        domain = list("CAPACITY", "UNMET DEMAND"),
        range = list("#6baed6", "#bdd7e7")
      ),
      legend = list(title = NULL, orient = "top-right")
    )
  )
)

vegawidget(as_vegaspec(spec_1d))
```

5)
```{r}
#| warning: false
#| message: false

spec_1e <- list(
  `$schema` = vega_schema(),
  data = list(values = long_data),
  layer = list(
    list(
      mark = list(type = "line", color = "#aec7d8"),
      encoding = list(
        x = list(
          field = "date",
          type = "nominal",
          axis = list(title = NULL, labelAngle = -45)
        ),
        y = list(
          field = "value",
          type = "quantitative",
          axis = list(title = NULL)
        ),
        detail = list(field = "date", type = "nominal")
      )
    ),
    list(
      mark = list(type = "point", size = 80),
      encoding = list(
        x = list(
          field = "date",
          type = "nominal"
        ),
        y = list(
          field = "value",
          type = "quantitative"
        ),
        fill = list(
          field = "type",
          type = "nominal",
          scale = list(
            domain = list("DEMAND", "CAPACITY"),
            range = list("white", "#6baed6")
          ),
          legend = list(title = NULL, orient = "top-right")
        ),
        stroke = list(value = "#6baed6")
      )
    )
  ),
  width = 500,
  height = 300
)

vegawidget(as_vegaspec(spec_1e))
```

6)
```{r}
#| warning: false
#| message: false

unmet_line_data <- data.frame(
  date = supply_demand$date,
  unmet = supply_demand$demand - supply_demand$capacity
)

spec_1f <- list(
  `$schema` = vega_schema(),
  data = list(values = unmet_line_data),
  layer = list(
    list(
      mark = list(type = "line", color = "#6baed6"),
      encoding = list(
        x = list(
          field = "date",
          type = "nominal",
          axis = list(title = NULL, labelAngle = -45)
        ),
        y = list(
          field = "unmet",
          type = "quantitative",
          axis = list(title = NULL)
        )
      )
    ),
    list(
      mark = list(type = "point", size = 80, filled = TRUE, color = "#6baed6"),
      encoding = list(
        x = list(
          field = "date",
          type = "nominal"
        ),
        y = list(
          field = "unmet",
          type = "quantitative"
        )
      )
    )
  ),
  width = 500,
  height = 300
)

vegawidget(as_vegaspec(spec_1f))
```

### Exercise 2

a) 
```{r}
#| warning: false
#| message: false

library(vegawidget)
library(jsonlite)

jobs_url <- "https://cdn.jsdelivr.net/npm/vega-datasets@2.8.0/data/jobs.json"
jobs <- fromJSON(jobs_url)

sum_perc <- aggregate(perc ~ year + sex, data = jobs, FUN = sum)

spec_2a <- list(
  `$schema` = vega_schema(),
  data = list(values = sum_perc),
  mark = list(type = "line", point = TRUE),
  width = 500,
  height = 300,
  encoding = list(
    x = list(
      field = "year",
      type = "quantitative",
      axis = list(title = "year", format = "d")
    ),
    y = list(
      field = "perc",
      type = "quantitative",
      axis = list(title = "sum of perc", format = ".0%")
    ),
    color = list(
      field = "sex",
      type = "nominal",
      legend = list(title = "sex")
    )
  )
)

vegawidget(as_vegaspec(spec_2a))
```

The sum of `perc` stays close to 1 (100%) for both men and women across every year. 
This means `perc` represents the share of workers of that sex in a given job — 
so all job percentages for men add up to 100%, and same for women separately.


b)
```{r}
#| warning: false
#| message: false

library(vegawidget)
library(jsonlite)

jobs_url <- "https://cdn.jsdelivr.net/npm/vega-datasets@2.8.0/data/jobs.json"
jobs <- fromJSON(jobs_url)

years_chosen <- c(1850, 1900, 1950, 2000)
jobs_filtered <- jobs[jobs$year %in% years_chosen, ]

spec_2b <- list(
  `$schema` = vega_schema(),
  data = list(values = jobs_filtered),
  spec = list(
    mark = list(type = "point"),
    width = 250,
    height = 250,
    transform = list(
      list(pivot = "sex", value = "perc", groupby = list("job", "year"))
    ),
    encoding = list(
      x = list(
        field = "men",
        type = "quantitative",
        axis = list(title = "men", format = ".0%")
      ),
      y = list(
        field = "women",
        type = "quantitative",
        axis = list(title = "women", format = ".0%")
      ),
      tooltip = list(
        list(field = "job"),
        list(field = "men", format = ".0%"),
        list(field = "women", format = ".0%")
      )
    )
  ),
  facet = list(
    field = "year",
    type = "ordinal",
    columns = 2
  )
)

vegawidget(as_vegaspec(spec_2b))
```

c)
```{r}
#| warning: false
#| message: false

library(dplyr)

jobs_url <- "https://cdn.jsdelivr.net/npm/vega-datasets@2.8.0/data/jobs.json"
jobs <- fromJSON(jobs_url)

years_chosen <- c(1850, 1900, 1950, 2000)
jobs_filtered <- jobs[jobs$year %in% years_chosen, ]

jobs_wide <- jobs_filtered %>%
  select(job, year, sex, count) %>%
  tidyr::pivot_wider(names_from = sex, values_from = count) %>%
  mutate(
    total = men + women,
    perc_men = men / total,
    perc_women = women / total
  )

spec_2c <- list(
  `$schema` = vega_schema(),
  data = list(values = jobs_wide),
  spec = list(
    mark = list(type = "point"),
    width = 250,
    height = 250,
    encoding = list(
      x = list(
        field = "perc_men",
        type = "quantitative",
        axis = list(title = "% men in job", format = ".0%")
      ),
      y = list(
        field = "perc_women",
        type = "quantitative",
        axis = list(title = "% women in job", format = ".0%")
      ),
      tooltip = list(
        list(field = "job"),
        list(field = "perc_men", format = ".0%"),
        list(field = "perc_women", format = ".0%")
      )
    )
  ),
  facet = list(
    field = "year",
    type = "ordinal",
    columns = 2
  )
)

vegawidget(as_vegaspec(spec_2c))
```

###Exercise 3

```{r}
#| warning: false
#| message: false

library(vegawidget)
library(tidyr)
library(dplyr)

gas <- read.csv("pump_price_for_gasoline_us_per_liter.csv", check.names = FALSE)
gas[, -1] <- lapply(gas[, -1], function(x) as.numeric(as.character(x)))

countries_url <- "https://raw.githubusercontent.com/lukes/ISO-3166-Countries-with-Regional-Codes/master/all/all.csv"
regions <- read.csv(countries_url, check.names = FALSE)
colnames(regions) <- make.names(colnames(regions))

gas_long <- gas %>%
  pivot_longer(-country, names_to = "year", values_to = "price_per_liter") %>%
  filter(!is.na(price_per_liter)) %>%
  mutate(
    year = as.numeric(year),
    price_per_gallon = price_per_liter * 3.785
  )

regions_clean <- regions %>%
  select(name, alpha.3, region) %>%
  rename(country = name, code = alpha.3)

gas_merged <- gas_long %>%
  left_join(regions_clean, by = "country")

unmatched <- gas_merged %>%
  filter(is.na(code)) %>%
  distinct(country)

gas_merged <- gas_merged %>% filter(!is.na(region), region != "")

spec_3 <- list(
  `$schema` = vega_schema(),
  data = list(values = gas_merged),
  mark = list(type = "line", opacity = 0.4),
  width = 500,
  height = 350,
  encoding = list(
    x = list(
      field = "year",
      type = "quantitative",
      axis = list(title = "year", format = "d")
    ),
    y = list(
      field = "price_per_gallon",
      type = "quantitative",
      axis = list(title = "price (USD per gallon)")
    ),
    color = list(
      field = "region",
      type = "nominal",
      legend = list(title = "region")
    ),
    detail = list(field = "country", type = "nominal")
  )
)

vegawidget(as_vegaspec(spec_3))

print(unmatched)
```

Yes, some countries in the gas price data failed to match with the regions dataset. 
This is because country names are spelled differently between the two datasets — 
for example "Russia" vs "Russian Federation", or "South Korea" vs "Korea, Republic of".











